---
title: "RANKL_Splicing"
output: html_document
date: "2025-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r, results='hide'}
library(SpliceWiz)
library(DESeq2)
library(tidyverse)
library(pheatmap)
```

## Build SpliceWiz Reference

```{r}
ref_path <- file.path('AltSplicing/splicewiz/ref')
fasta_path = 'C:/CassatLab/STm_RANKL_RNASeq/GRCm39/Mus_musculus.GRCm39.dna_rm.primary_assembly.fa'
gtf_path = 'C:/CassatLab/STm_RANKL_RNASeq/GRCm39/Mus_musculus.GRCm39.113.gtf'

buildRef(
    reference_path = ref_path,
    fasta = fasta_path,
    gtf = gtf_path,
    genome_type="mm10"
)
```

## Process BAM Files

```{r}
align_directory<-"C:/CassatLab/STm_RANKL_RNASeq/02_Mapping/aligned"
bams <- findBAMS(align_directory, level = 0)

pb_path <- file.path('AltSplicing/splicewiz/pb_output')
processBAM(
    bamfiles = bams$path,
    sample_names = bams$sample,
    reference_path = ref_path,
    output_path = pb_path,
    n_threads = 8
    )
```

## Collate Data

```{r}
expr <- findSpliceWizOutput(pb_path)

nxtse_path <- file.path('AltSplicing/splicewiz/NxtSE_output')
collateData(
    Experiment = expr,
    reference_path = ref_path,
    output_path = nxtse_path,
    n_threads = 8
)
```

```{r}
expSetup <- read.csv("seq_reference_splice.txt", sep="\t")

expr <- merge(expr,expSetup,by="sample",all.x=TRUE) %>%
  as.data.frame()
```


```{r}
se <- makeSE(nxtse_path)
```

```{r}
colData(se)$condition <- expr$condition
```

## Filter Low Confidence Alternative Splice Events

```{r}
se.filtered <- se[applyFilters(se),]
```

## Differential Alt. Splicing Analysis with DESeq2

```{r}
res_DESeq <- ASE_DESeq(
    se = se.filtered,
    test_factor = "condition",
    test_nom = "4h_STm_100",
    test_denom = "4h_STm_0",
    n_threads = 8
)
```

## Filter out low delta-PSI (change in "percent spliced in") events.

```{r}
res_DESeq_Filt <- res_DESeq[res_DESeq$abs_deltaPSI > 0.05,]
```

## Plot Volcano Plot for Given Condition

```{r}
ggplot(res_DESeq_Filt,
        aes(x = log2FoldChange, y = -log10(padj))) + 
    geom_point() +
    labs(title = "Differential analysis - 4h_STm_100 vs 4h_STm_0",
         x = "Log2-fold change", y = "Adj. P-value (-log10)")
```

```{r}
# Set up list of comparisons, numerator first, denominator second
comp_list = list(
  list("4h_STm_100", "4h_STm_0"),
  list("4h_STm_0", "4h_mock_0"),
  list("4h_mock_100", "4h_mock_0")
)

for (x in comp_list) {

  # Run DESeq 2 on Comparison of Interest
  res_DESeq <- ASE_DESeq(
      se = se.filtered,
      test_factor = "condition",
      test_nom = x[[1]],
      test_denom = x[[2]],
      n_threads = 8
  )
  
  # Filter to splice events with delta-PSI > 5% 
  res_DESeq_Filt <- res_DESeq[res_DESeq$abs_deltaPSI > 0.05,]
  
  # Save to file for later interpretation
  write.csv(res_DESeq_Filt, file=paste0("AltSplicing/",x[[1]],"_vs_",x[[2]],".csv"), row.names = FALSE)

}
```

## Setup Gene Ontology

```{r}
onto_ref_path <- file.path('AltSplicing/splicewiz/ref')
ontology <- viewGO(onto_ref_path)
head(ontology)
```

```{r}
go_byASE <- goASE(
  enrichedEventNames = res_DESeq_Filt$EventName,
  universeEventNames = res_DESeq$EventName,
  se = se
)
head(go_byASE)
```

```{r}
plotGO(go_byASE, filter_n_terms = 12)
```

## Heatmap Generation

First select the top ~100 alternative splicing events from each dataset.
I'll do this by iterating through the saved files, collecting the top __ results (when adjusted for p-val), and removing duplicates at the end.

```{r}
# Set up list of comparisons, numerator first, denominator second
comp_list = list(
  list("4h_STm_100", "4h_STm_0"),
  list("4h_STm_0", "4h_mock_0"),
  list("4h_mock_100", "4h_mock_0")
)

full_plot_df = data.frame(matrix(ncol = 26, nrow = 0))

for (x in comp_list) {
  
  res_temp = read.csv(paste0("AltSplicing/",x[[1]],"_vs_",x[[2]],".csv"))
  
  subset <- res_temp %>%
    arrange(desc(abs_deltaPSI)) %>%  # arrange in descending order
    subset(padj <= 1e-5) %>% 
    slice(1:100)
  
  names(full_plot_df) <- names(subset) 
  
  full_plot_df <- rbind(full_plot_df, subset)
}

full_plot_df <- full_plot_df %>% 
  distinct(EventName)

```




```{r}
mat <- makeMatrix(
    se.filtered,
    event_list = full_plot_df$EventName,
    method = "PSI"
)
```

```{r}
anno_col_df <- as.data.frame(colData(se.filtered)) %>%
    arrange(condition)
anno_col_df <- anno_col_df[, 1, drop=FALSE]

pheatmap(mat, annotation_col = anno_col_df,
         cluster_cols = TRUE,
         show_rownames = TRUE, 
         show_colnames = FALSE,
         treeheight_col = 15,
         treeheight_row = 25,
         scale='none',
         cellheight=10,
         cellwidth=20,
         angle_col=315,
         main = "Log2 Transformed Heatmap \n Alternative Splicing Events by % Spliced In",
         filename="AltSplicing/test_heatmap.png")
```

